Ownership types in theory and practice (in Rust)

### 2023-01-17 paper/paper.tex
Added example for owners-as-dominators

@@

\begin{figure}
  \begin{tikzpicture}[object/.style={rectangle, draw, very thick, minimum size=20}]
    \node[object] (World) {world};
    \node[object] (A) [below=of World] {a};
    \node[object] (C) [right=of A] {c};
    \node[object] (B) [below=of A] {b};

    \draw[->,thick] (World) to node[left] {owns} (A);
    \draw[->,thick] (World) to node[right] {owns} (C);
    \draw[->,thick] (A) to node[left] {owns} (B);
    \draw[->,thick,dashed] (C) to node[right] {invalid access} (B);
  \end{tikzpicture}

  \caption{Owners-as-modifiers}
  \label{fig:owners-as-modifiers}
\end{figure}

    \draw[->,thick] (World) to (Car);
    \draw[->,thick] (World) to (Driver);
    \draw[->,thick] (Car) to (Engine);
    \draw[->,thick,dashed] (Car) to node[above] {valid} (Driver);
    \draw[->,thick,dashed] (World) edge[bend right=45] node[left] {invalid} (Engine);
  \end{tikzpicture}

  \caption{Owners-as-modifiers \cite{flexible-alias-protection}, solid lines indicate "owns"}
  \label{fig:owners-as-modifiers2}
\end{figure}

@@

An implication of this model is that any external reference to an object is only possible through its owner.
Figure \ref{fig:owners-as-modifiers} provides an example of an invalid reference from object \emph{c} to \emph{b}, as there exists no direct link in the ownership hierarchy.
% TODO: what does this implicate further?

It is the strictest discipline.
The owners-as-dominators principle provides a simple, clear and strong guarantee but makes programming more difficult by not allowing common idioms through aliasing.
% TODO: example of such idiom that isn't representable

% IMPORTANT TODO: question: can an object access owned children transitively?

% Advantages: useful for reasoning about properties of code.

>>>>

    \draw[->,thick] (World) to (Car);
    \draw[->,thick] (World) to (Driver);
    \draw[->,thick] (Car) to (Engine);

    % TODO: make green darker with:
    % \definecolor{mynicegreen}{RGB}{102,252,102}

    \draw[->,thick,dashed,green] (Car) edge[bend left] node[right] {owner} (Engine);
    % \draw[->,thick,dashed,green] (Engine) edge[bend left] node[left] {inside} (Car);
    \draw[->,thick,dashed,green] (Engine) edge[bend left=25] node[left] {inside} (World);
    \draw[->,thick,dashed,green] (Car) edge node[above] {sibling} (Driver);
    \draw[->,thick,dashed,red] (World) edge[bend right=90] node[left] {invalid} (Engine);
    \draw[->,thick,dashed,red] (Driver) edge[bend left] node[right] {invalid} (Engine);
  \end{tikzpicture}

  \caption{Owners-as-dominators \cite{flexible-alias-protection}: Solid lines indicate "owns" and dotted lines indicate "references".}
  \label{fig:owners-as-dominators}
\end{figure}

Figure \ref{fig:owners-as-dominators} was adapted from an example by Clarke et al. \cite{flexible-alias-protection} and demonstrates valid (green) and an invalid (red) references in an ownership hierarchy.
% TODO: make more precise
The validity of each valid reference is proven by each of the rules defined above.

\begin{enumerate}
  \item The object \emph{car} owns the object \emph{engine}, meaning that the former can validly reference the latter according to the first rule.
  \item The objects \emph{car} and \emph{driver} are siblings, meaning that they can validly reference each other according to the second rule.
  \item The object \emph{world} is outside the object \emph{engine} because \emph{engine} is transitively owned by \emph{world} via \emph{car}.
  Therefore, according to the third rule, \emph{engine} can validly reference \emph{world}.
\end{enumerate}

@@

An implication of this model is that any external reference to an object is only possible through its owner.
This is also examplified in Figure \ref{fig:owners-as-dominators} through the object \emph{world} trying to access \emph{engine}.
This is not allowed in the owners-as-dominators model.
Instead, \emph{world} can only reference \emph{engine} through the \emph{engine}'s owner \emph{car}.

Its strictness can be seen as both an advantage and disadvantage of the owners-as-dominators model.
% TODO: reference ownership survey better, as this is nearly copied
The advantage is that model provides a simple, clear and strong guarantee that allows for reasoning about various properties of the code.
% TODO: provide example of advantages
However, it also makes programming more difficult by not allowing common idioms that involve aliasing.
% TODO: example of such idiom that isn't representable

@@

% IMPORTANT TODO: question: can an object access owned children transitively? How is it in Rust (if someone asks me this question)?
